---
title: "Halloween candy: analysis report"
output: html_notebook
---

```{r}
library(tidyverse)
library(readxl)
library(janitor)
library(here)
```

# Load and inspect data

```{r}
excel_sheets(here("raw_data/boing-boing-candy-2015.xlsx"))
excel_sheets(here("raw_data/boing-boing-candy-2016.xlsx"))
excel_sheets(here("raw_data/boing-boing-candy-2017.xlsx"))
```

Each excel has only one sheet, each excel file is a dataset for one year. 

Read these in:
```{r}
x2015 <- read_excel(here("raw_data/boing-boing-candy-2015.xlsx"), sheet = "Form Responses 1")
x2016 <- read_excel(here("raw_data/boing-boing-candy-2016.xlsx"), sheet = "Form Responses 1")
x2017 <- read_excel(here("raw_data/boing-boing-candy-2017.xlsx"), sheet = "responses (2) (1).csv")
```


# Analysis questions

Note: more info on the data source [here](https://www.scq.ubc.ca/so-much-candy-data-seriously/).

1. What is the total number of candy ratings given across the three years. (Number of candy ratings, not the number of raters. Don’t count missing values)
2. What was the average age of people who are going out trick or treating?
3. What was the average age of people who are not going trick or treating?
4. For each of joy, despair and meh, which candy bar received the most of these ratings?
5. How many people rated Starburst as despair?

For the next three questions, count despair as -1, joy as +1, and meh as 0.
6. What was the most popular candy bar by this rating system for each gender in the dataset ?
7. What was the most popular candy bar in each year?
8. What was the most popular candy bar by this rating for people in US, Canada, UK, and all other countries?

Which variables do we need to answer the analysis questions? We need:
* All candy ratings (1 rating of 1 candy bar by 1 rater in 1 year = 1 observation)
* Age of rater, where available
* Whether or not the rater goes out trick or treating
* To be able to summarise count of ratings (joy, meh, despair) for each candy bar
* Gender of rater, where available
* To be able to filter for Starburst and count despair ratings
* To be able to convert ratings to +1, 0, -1 and sum across these 3 (new column)
* To be able to group by year
* To be able to group by country

# Approach for joining data (and making it 'tidy' long format)

Step 1: Make tidy data: 
  * We want variables: rater_info (ToT_activity, age, gender, country), rating_year, type of candy bar <chr>, rating (joy, meh, despair) 
  * Make sure variables are in same (and desired) data type in all 3 dfs
  * add a column for year in each dataset
  * add column with NAs for any of the above variables where data does not exist in that year's df 

Step 2: Join the tidy dataframes (append rows)
* Try tidying each sheet first, then joining all 3 together. Then deal with inconsistent candy bar names in later cleaning step, i.e.:
  * Make matching variables and rated objects (candy bars) consistent across the 3 dataframes
  
Aim: I want to make one combined df. that looks like the following, whereby 1 rating of 1 candy item by 1 rater in 1 year = 1 observation:

| year <dttm( or int(YYYY)> | id <int> | age <int> | gender <chr> | country <chr> | goes_trick_or_treating <chr or lgl> | candy_item <chr> | rating <chr>  | [And later: rating_value <dbl>] |
|----|----|----|----|----|----|----|----|----|
| 2015 | 5058 | 23 | Male | USA | TRUE | Twix | JOY | 1 |
| 2016 | 1123 | 36 | Female | Canada | FALSE | Twix | DESPAIR | -1 |
| 2016 | 1123 | 36 | Female | Canada | FALSE | Trail mix | MEH | 0 |

## Variables, rows and data types

Here, I have inspected the variables within each year's dataset, and decided what to remove, what to keep and how to keep it, to achieve the above aim of combined tidy data.

### 2015

```{r}
ncol(x2015) # 124 columns
colnames(x2015)
```

```{r}
glimpse(x2015) # 5630 rows
# data types: mostly <chr>, some <lgl> at end, 1 <dttm> at beginning
```

Variables to keep/convert:

| Column # | Current variable name <type> | Action required to make tidy data |
| ----- | ----- | ----- |
| 1 | `Timestamp` <dttm> | Convert to `year` <dttm (YYYY)> |
| 2 | `How old are you?` <chr> | Convert to `age` <int> |
| 3 | `Are you going actually going trick or treating yourself?` <chr> |  Convert to `goes_trick_or_treating` <chr or lgl> |
| 4-17,19-22,24-25,29-32,35-37,39-40,42-44,46-55,57-81,83-89,91-92,96,114,115 | `Candy or other substance` <chr> | Remove non-candy items (see below) then make tidy -> `Candy_item` <chr> + `Rating` <chr> |
| 97,100-113,116-124 | Miscellaneous  | Remove |
| 98 | `Please list any items not included above that give you JOY.` | Extract answers --> new cols with JOY as value, and include in tidying step |
| 99 | `Please list any items not included above that give you DESPAIR.` | Extract answers --> new cols with DESPAIR as value, and include in tidying step |

Decisions on variables to keep/remove:

* Some are not candy (either obviously or by searching online), so removed:
  * Other food items, not candy: [34] `Healthy fruit`, [41] `Kale smoothie`, [56] `Minibags of chips` (crisps, not candy), [82] `Spotted Dick` (cake, not candy), [90] `Peanut butter jars` (nut butter, not candy), [94] `White bread`, [95] `Whole wheat anything`
  * Pharmaceuticals: [26] `[Generic Brand Acetaminophen]`, [93] `[Vicodin]`
  * other items, not food: [18] `Cash...`, [23] `Dental paraphenalia`, [27] `Glow sticks`, [28] `Broken glow stick`, [33] `Creepy Religious comics...`, [38] `Hugs...`, [45] `Lapel pins`
* Items kept in include candy that I recognise by name or have identified as candy by searching online, for example:
  * [63] `Pencils`, [88] `Peterson Brand sidewalk chalk`, [91] `Trail mix` - note decided to keep this in as it is a high sugar content snack.

```{r 2015-subset}
# keep wanted columns, and put 98-99 (additional ratings) at end
x2015_subset <- x2015 %>% 
  select(1:17,19:22,24,25,29:32,35:37,39:40,42:44,46:55,57:81,83:89,91,92,96,114,115,98,99)

head(x2015_subset) # 84 variables

# rename variables about the rater & rating year
# and reposition to end with: 
# [1]year [2]age [3]gender [4]country [5]goes_trick_or_treating
x2015_subset_converted_all <- x2015_subset %>% 
  # make an id column to retain unique identifier for each person
  mutate(id = c(1:5630), .before = "Timestamp") %>%
  rename(age = "How old are you?",
          goes_trick_or_treating = "Are you going actually going trick or treating yourself?",
          year = "Timestamp",
          additional_joy = "Please list any items not included above that give you JOY.",
          additional_despair = "Please list any items not included above that give you DESPAIR.") %>%
  # recode some specific age values before converting to numeric (see "Coerced age data" below)
  mutate(age = case_when(
    id == 378 ~ "45",
    id == 792 ~ "37",
    id == 1210 ~ "43",
    id == 1571 ~ "46",
    id == 1629 ~ "40",
    id == 2207 ~ "37",
    id == 2934 ~ "50",
    id == 3384 ~ "27",
    id == 3626 ~ "50",
    id == 4798 ~ "42",
    id == 5624 ~ "50",
    .default = age)) %>% 
  mutate(age = as.integer(age), # Note: this introduces NAs by coercion
         year = as.numeric(format(year, "%Y"))) %>% 
  # make empty columns for gender and country (missing in this year's dataset)
  mutate(gender = rep(NA_character_, nrow(x2015_subset)), .after = age) %>% 
  mutate(country = rep(NA_character_, nrow(x2015_subset)), .after = gender) %>% 
  filter(!id == 1573) # remove 1 duplicate 
  # (see "Coerced age data" below for reasoning)
  
# check NAs in age
x2015_subset_converted_all %>% 
  filter(is.na(age)) # 273 rows, which is 285 (original NAs) minus 11 recoded minus 1 duplicate

# If processing additional_joy/despair ratings, do this here using df: `..._all`
# For now, proceed without these additional ratings, remove these columns:

x2015_subset_converted_sub <- x2015_subset_converted_all %>% 
  select(-c(additional_joy,additional_despair))

dim(x2015_subset_converted_sub) # 5630 observations x 85 variables
head(x2015_subset_converted_sub) # cols 1:6 are rater_info, 7-85 are candy items, 85-86

# 'tidy' subsetted data to long format with
# one column for candy items [7-85], one column for rating
# expect 5629 x 85 to become 
# (5629 raters * 79 items) x 8 = 444,691 x 8
x2015_tidy <- x2015_subset_converted_sub %>% 
  pivot_longer(cols = -c(year, id, age, gender, country, goes_trick_or_treating),
               names_to = "candy_item",
               values_to = "rating"
               )
dim(x2015_tidy) # 444,691 obs x 8 var - as expected
head(x2015_tidy, n = 10)
tail(x2015_tidy, n = 10)
```
Note: this tidy data contains ratings from 5,629 individual raters (with ids 1-5630; excluding 1573, duplicate entry) about 79 different candy items.

Currently, this does not include additional JOY and DESPAIR ratings given as free text entries to 2 questions in the raw data.

Note that goes_trick_or_treating is <chr> type.
Attempted mutate(goes_trick_or_treating = as.logical(goes_trick_or_treating)) =  this coerces to NA, so need different approach to make T/F  values - not done yet.

The above code that is needed to generate the tidy version of 2015 data has been included in `cleaning.R` script; the code to check results has not been included, and neither has code to include the additional ratings columns.

#### Coerced age data

Warning: converting age data from <chr> to numeric coerces some values into NAs. Inspect raw data to understand what I might be missing by doing this.

```{r}
# check for NAs in converted age column
x2015_id_na_age <- x2015_subset_converted_sub %>% 
  filter(is.na(age)) %>% # 285 rows
  # make vector of IDs to filter for to inspect raw data (with IDs)
  select(id) %>% 
  pull()
   
x2015_id_age_raw <- x2015 %>% 
  # add same ids to raw data
  mutate(id = c(1:5630), .before = Timestamp) %>% 
  rename(age_raw = "How old are you?") %>% 
  # filter for id's with NA in converted age
  filter(id %in% x2015_id_na_age) %>% 
  select(id, age_raw) %>% # several NAs, filter out 
  filter(!is.na(age_raw)) # 85 answers

# test how to recode these - reuse this code in cleaning script (as above)
x2015_id_age_raw %>% 
  mutate(age = case_when(
    id == 378 ~ "45",
    id == 792 ~ "37",
    id == 1210 ~ "43",
    id == 1571 ~ "46",
    id == 1629 ~ "40",
    id == 2207 ~ "37",
    id == 2934 ~ "50",
    id == 3384 ~ "27",
    id == 3626 ~ "50",
    id == 4798 ~ "42",
    id == 5624 ~ "50",
    .default = age_raw
  )) %>% 
  mutate(age_num = as.integer(age)) %>% 
  group_by(age_num) %>% 
  summarise(count = n()) # recoded 11 values, 74 NAs remaining
```

Inspecting these 85 answers, there are a few I will recode to include in dataset, the rest are not meaningful enough and can be coerced to NAs. 

To recode to include:

| id | age_raw | recode as age: |
|-----|-----|-----|
| 378 | "45, but the 8-year-old Huntress..." | 45 |
| 792 | "37 (I'm taking a child)" | 37 |
| 1210 | "Good Lord!  I'm 43!" | 43 |
| 1571 | "46:" | 46 |
| 1573 | "46:" | 46 |
| 1629 | "40. Deal with it." | 40 |
| 2207 | "37," | 37 |
| 2934 | "50 (despair)" | 50 |
| 3384 | "27^" | 27 |
| 3626 | "50, taking a 13 year old." | 50 |
| 4798 | "42 - I'm taking my kid" | 42 |
| 5624 | "50t" | 50 |

Notes:

* Do not just remove punctuation, there are some entries that are "30+" or "50+" which cannot be recoded as an age.
* 1571 and 1573 are exactly the same - suspicious there may be duplicate entries. Checked full data for these IDs (see below) and decided to remove 1573 from the dataset, since it is a duplicate entry.
* Recode age values for the IDs as above (excluding 1573, to be removed) =  expect to end up with (285 - 11) = 274 NAs in final data.

```{r}
# check if 1571 and 1573 are duplicate entries

x2015_subset_converted_sub %>% 
  filter(id %in% c(1571,1573)) # exactly the same entries

x2015 %>% 
  mutate(id = c(1:5630), .before = Timestamp) %>% 
  filter(id %in% c(1571,1573))
# Timestamp: entered <1 minute apart
# all answers, including free text fields, are exactly the same
# consider these duplicates and remove 1573
  
```

#### Additional data not yet in columns

Columns 98-99 in x2015 (raw data) are free text fields used to record additional items that are rated as JOY or DESPAIR, respectively, by the rater.

If time, I could add these additional data into the 2015 dataset before making it long format. For now, given the number of entries here (see below: 1000s additional rows to process), I will proceed without using this data, and thus discard it from the data to proceed with.

```{r}
# Create separate table for additional items rated as JOY or DESPAIR with person id
x2015_additional_ratings <- x2015_subset_converted_all %>% 
  select(id, additional_joy, additional_despair)
head(x2015_additional_ratings, n=10)

# see list of additional candy items rated JOY
x2015_additional_joy <- x2015_additional_ratings %>% 
  filter(!is.na(additional_joy)) %>% 
  select(id, additional_joy) #2436 rows

# see list of additional candy items rated DESPAIR
x2015_additional_despair <- x2015_additional_ratings %>% 
  filter(!is.na(additional_despair)) %>% 
  select(id, additional_despair) #1775 rows
```

#### Attempt analyses

Try out some analysis Qs on 2015 data, where possible, to check cleaning steps and note what else is required.

##### Total number of ratings in 2015

(Q1: What is the total number of candy ratings given across the three years.)

```{r}
# total number of ratings in 2015
x2015_tidy %>% 
  filter(!is.na(rating)) %>% 
  group_by(rating) %>% 
  summarise(count = n()) %>% 
  summarise(total_count = sum(count)) %>% 
  pull()
```

Note the ratings include only JOY, DESPAIR, or NA (where no rating given for a candy bar). There is no MEH in 2015.

##### Candy with the most JOY ratings

(Q4: For each of joy, despair and meh, which candy bar received the most of these ratings?)

```{r}
# Find candy item with most JOY ratings
x2015_tidy %>% 
  filter(rating == "JOY") %>% 
  group_by(candy_item) %>% 
  summarise(count = n()) %>% 
  # arrange(desc(count)) %>% 
  slice_max(count, n = 2) # include 2nd place as 1st place is generic
```

#### Summary from tidying 2015

* Not yet cleaned within columns: countries, candy_items, in particular for 2015. Waiting to have full df with all three years, to do this in one go.
* Not yet added in additional candy_items with JOY/DESPAIR ratings, stored in `x2015_additional_ratings` (subsetted into `x2015_additional_joy` and `x2015_additional_despair`).
* Any analyses by age or gender will filter out 2015 data because all _NA_ here.
* Written 2 forms of csv - one wide format for looking at average ages, 1 row = 1 person, one long format for looking at counts of ratings

```{r}
# read in csv written from cleaning script
tidy_2015 <- read_csv(here("clean_data/2015_clean_long.csv"))
head(tidy_2015)
```


### 2016

```{r}
ncol(x2016) # 123 columns
colnames_2016 <- as.data.frame(colnames(x2016))
```

```{r}
glimpse(x2016) # 1259 rows
# data types: mostly <chr>, 1 <lgl> at end, 1 <dttm> at beginning
```

#### 2016 cleaning

Follow 2015 cleaning steps, see what needs changing:

* 1. subset to remove unwanted columns
* 2. clean up included variables --> keep as clean_wide data
* 3. pivot to long format for tidy data --> keep as clean_long data

#### 1. subset to remove unwanted columns

Decisions to change or remove columns:

| Column # | Current variable name <type> | Action & reasoning |
| ----- | ----- | ----- |
| 1 | `Timestamp` <dttm> | Convert to `year` <dttm (YYYY)> |
| 2 | `Are you going actually going trick or treating yourself?` <chr> |  Convert to `goes_trick_or_treating` <chr> |
| 3 | `Your gender:` <chr> |  Convert to `gender` <chr> |
| 4 | `How old are you?` <chr> | Convert to `age` <int> |
| 5 | `Which country do you live in?` <chr> |  Convert to `country` <chr> |
| 6 | `Which state, province, county do you live in?` <chr> |  Remove, not needed |
| 7-106 | Candy and other items | Remove non-candy items, as noted below, then make tidy -> `Candy_item` <chr> + `Rating` <chr> |
| 107-108 | Additional JOY and DESPAIR-rated items | Process and include, if time (as per 2015) |
| 109 | `Please leave any witty, snarky or thoughtful remarks or comments regarding your choices.` | remove for now; if time, inspect to see if additional data worth keeping |
| 110-123 | Miscellaneous | Remove, irrelevant |

Decisions on variables to keep/remove:

* Non-candy items to remove:
  * Other food/drink items, not candy: [22] `Chardonnay`, [38] `Healthy fruit`, [49] `Kale smoothie`, [69] `Minibags of chips`, [90] `Spotted Dick`, [104] `White bread`, [105] `Whole wheat anything`
  * Pharmaceuticals: [31] `[Generic Brand Acetaminophen]`, [102] `[Vicodin]`
  * Other items, not food: [12] `Bonkers (the board game)`, [15] `Broken glow stick`, [21] `Cash...`, [26] `Creepy Religious comics...`, [27] `Dental paraphenalia`,
[32] `Glow sticks`, [43] `Hugs...`, [79] `Person of interest...` (DVD)
* Some new items retained as candy (beyond 2015 items):
  * [94] `Sweetums` - from fictional TV series, but online search suggests some candies made under this name at some point, so possible they could have been handed out at Halloween
  * [103] `Whatchamacallit Bars` - a chocolate-based baked item

```{r}
## step 1: subset to remove unwanted columns
x2016_subset <- x2016 %>% 
  select(1:11,13,14,16:20,23:25,28:30,33:37,39:42,44:48,50:68,
         70:78,80:89,91:101,103,106) 
    # not yet including 107:109 for additional ratings and comments
# 1259 obs x 89 var
```


#### 2. clean up included variables

First, check which age data values will be coerced to NA by converting to numeric (learning from 2015 cleaning steps):

```{r}
x2016_subset %>% 
  mutate(id = c(1:nrow(x2016_subset)), .before = "Timestamp") %>% 
  rename(age_raw = "How old are you?") %>% 
  mutate(age = as.integer(age_raw)) %>% 
  filter(is.na(age)) %>% 
  select(id, age_raw, age)
```
68 rows were coerced to NA. By inspection, none are worth recoding before converting age to numeric.

```{r}
## step 2: clean up included variables to produce required common df structure: 
# [1]id [2]year [3]age [4]gender [5]country 
# [6]goes_trick_or_treating [7]candy_item [8]rating
x2016_subset_converted <- x2016_subset %>% 
  # make an id column to retain unique identifier for each person
  mutate(id = c(1:nrow(x2016_subset)), .before = "Timestamp") %>% 
  # rename variables
  rename(year = "Timestamp",
         goes_trick_or_treating = "Are you going actually going trick or treating yourself?",
         gender = "Your gender:",
         age = "How old are you?",
         country = "Which country do you live in?"
         ) %>%
  # change format of age and year data
  mutate(age = as.integer(age), # Note: this introduces NAs by coercion
         year = as.numeric(format(year, "%Y")))
```

```{r}
# inspect converted df
dim(x2016_subset_converted) # 1259 obs. x 90 var. 1 more col than subset: + id
head(x2016_subset_converted)
```

#### 3. pivot to long format for tidy data

```{r}
## step 3: make long format tidy data
# with one column for candy items 7:90, one column for rating
# expect 1259 x 90 df to become 
# (1259 raters * 84 items) x 8 = 105,756 x 8
x2016_tidy <- x2016_subset_converted %>% 
  pivot_longer(cols = -c(year, id, age, gender, country, goes_trick_or_treating),
               names_to = "candy_item",
               values_to = "rating")

dim(x2016_tidy) # 105,756 x 8
```

#### Summary

Steps worked, with minimal changes or additional steps compared to 2015. Note 2016 data includes country and gender, not yet inspected for cleaning (to do once all years combined).

### 2017

```{r}
ncol(x2017) # 120 columns
colnames(x2017)
```

```{r}
glimpse(x2017) # 2460 rows
# data types: mostly <chr>, some <dbl>, no <lgl>
```



## Check consistency of candy items in each year

```{r}
# list candy items in 2015 data
x2015_tidy %>% 
  distinct(candy_item)
```


# Inspect data within

* find missing values, NAs
* clean up country column - need USA, Canada, UK, all other countries
  * note in 2016 there is age data in the country column!
* clean up any other data, e.g. character

