---
title: "Data exploration"
output: html_notebook
---

```{r}
library(tidyverse)
library(readxl)
library(janitor)
```
# Load in raw data

```{r}
sheet_names <- excel_sheets("data/seabirds.xls")
```

```{r}
ship_data <- read_excel("data/seabirds.xls", sheet = "Ship data by record ID")
bird_data <- read_excel("data/seabirds.xls", sheet = "Bird data by record ID")
ship_codes <- read_excel("data/seabirds.xls", sheet = "Ship data codes")
bird_codes <- read_excel("data/seabirds.xls", sheet = "Bird data codes")
```

Note the error this flagged - it may mean that there is a column that contains mixed type data, or that data has spilled across columns. Tbc when I inspect the subset of data I need for this analysis.


# Analysis planning

## Requested insights

* Which bird had the most individual sightings?
* Which bird had the highest total count?
* Which bird had the highest total count above a latitude of -30?
* How many different types of birds were only ever seen in groups of 1?
* How many penguins were seen? (Hint: there are many types of penguin)

For each question where we ask for bird names, give the bird’s common name, scientific name and species abbreviation.

## Gather required data

1. Join ship_data and bird_data by `RECORD ID` - this is the common key that links the ship data (date of sightings, from which ship, at which lat/long position, who sighted, weather conditions, etc) with the bird data (which bird sighted, how many counted in the sighting, etc). I have decided to left join, to keep all the bird data, and only match ship data where there is a match (I don't need any ship records that don't match up to recorded bird sightings).

2. Retain only the following columns (for the above analysis questions) & rename for tidy data, to create a dataframe like this:

| record_id | ship_lat | common_name | sci_name | species_abbrv | count |

```{r}
bird_counts <-  left_join(bird_data, ship_data, by = "RECORD ID") # expect all 49,019 bird observations

bird_counts <- bird_counts %>% 
  select(`RECORD ID`, LAT, `Species common name (taxon [AGE / SEX / PLUMAGE PHASE])`, `Species  scientific name (taxon [AGE /SEX /  PLUMAGE PHASE])`, `Species abbreviation`, COUNT) %>% 
  clean_names() %>% 
  rename(
   common_name = species_common_name_taxon_age_sex_plumage_phase,
   sci_name = species_scientific_name_taxon_age_sex_plumage_phase
  )

glimpse(bird_counts)

# write subset of raw data to new csv file to continue with in next cleaning script
write_csv(bird_counts, "data/bird_counts_raw.csv")
```
* Note that text columns are character type, numeric columns are <dbl>. This means any NAs or missing values in numeric columns are standard NAs, blanks or numbers - i.e. no entries written in as 'none' for example.

```{r}
bird_counts
```


## Cleaning tasks: initial plan

* Numeric data:
  * check min and max make sense for latitudes and counts

* Common name: 
  * remove 'sensu lato' =  latin for "in the broad sense"
  * remove age codes: AD, JUV, IMM, 
  * remove plummage codes: PL# 
  * rename ...unidentified as unidentified - or check if better to keep as is and make new col for unidentified, to exclude by
  * Note some bird1 / bird 2 -- could rename as unidentified since not sure which one
  * Do distinct() to see if any more need recording
  * NO BIRDS RECORDED - note as an NA for bird count
  
* Scientific name:
  * Have age and plummage codes in this data too (as well as in species abbrev) - does it mean something? Check domain expertise
  * Some NAs
  * Several with "/" - unidentified / not sure which. Decide whether to keep.
  
  
  
  