---
title: "Analysis of seabird sightings"
output: html_notebook
---

```{r}
bird_counts <- read_csv("clean_data/bird_counts_raw.csv")

glimpse(bird_counts)
```

## Most individual sightings

_Task: Which bird had the most individual sightings? Give the bird’s common name, scientific name and species abbreviation._

Look for most frequent bird, where count = 1.

```{r}
bird_counts %>% 
  filter(count == 1) %>%  # 20,357 individual sightings
  group_by(species_abbreviation) %>% 
  summarise(lone_sightings = n()) %>% 
  arrange(desc(lone_sightings)) # 298 rows
```

Without recoding / cleaning bird names, there are 298 rows in the summary table when grouping by common_name, some of which are the same bird with different elements in their common name (such as plummage, age). When grouping by species_abbreviation, there are 297 (also contain plummage and age in the column), so this does not improve matters.

So, I do need to clean up these names to get a true total count per bird.

```{r}
# after cleaning common_name column:
bird_counts_cleaned %>% 
  filter(count == 1) %>%  # 20,357 individual sightings
  group_by(common_name, sci_name, species_abbreviation_clean) %>% 
  summarise(lone_sightings = n()) %>% 
  arrange(desc(lone_sightings)) 
```
If group by common_name only, now 150 rows, some with common_names including various different possible birds (separated by / ). However, wandering albatross has the most individual sightings by far, so I assume that this answer won't change even if I try to deal with the not-quite-specified birds.

If group by all 3 name columns, 202 rows, which indicates common name : sci name : species abbreviation is not 1:1:1 relationship. In error: Adelie Penguin might have multiple matches.



Also, the cleaning step worked, so I will write a function for this, to repeat over all 3 name columns

## Highest total count of sightings

_Task: Which bird had the highest total count? Give the bird’s common name, scientific name and species abbreviation._



## Birds sighted alone

_Task: How many different types of birds were only ever seen in groups of 1?_



## Penguin sightings

_Task: How many penguins were seen? (Hint: there are many types of penguin)_



